---
title: "Gene Usage"
date: '`r format(Sys.Date(), "%B %d %Y")`'
output:
  rmarkdown::html_vignette:
    toc_float: true
    toc_depth: 2
    vignette: >
      %\VignetteIndexEntry{Gene Usage}
      %\VignetteEncoding{UTF-8}
      %\VignetteEngine{knitr::rmarkdown}
---

```{r "setup", include = FALSE}
# Chunk opts
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 8,
  fig.height = 3
)
```

<br>

This vignette provides detailed examples for calculating and visualizing V(D)J gene usage. For the examples shown below, we use data for splenocytes from BL6 and MD4 mice collected using the 10X Genomics scRNA-seq platform. MD4 B cells are monoclonal and specifically bind hen egg lysozyme.

```{r}
library(djvdj)
library(Seurat)
library(ggplot2)
library(RColorBrewer)

# Load GEX data
data_dir <- system.file("extdata/splen", package = "djvdj")

gex_dirs <- c(
  BL6 = file.path(data_dir, "BL6_GEX/filtered_feature_bc_matrix"),
  MD4 = file.path(data_dir, "MD4_GEX/filtered_feature_bc_matrix")
)

so <- gex_dirs |>
  Read10X() |>
  CreateSeuratObject() |>
  AddMetaData(splen_meta)

# Add V(D)J data to object
vdj_dirs <- c(
  BL6 = system.file("extdata/splen/BL6_BCR", package = "djvdj"),
  MD4 = system.file("extdata/splen/MD4_BCR", package = "djvdj")
)

so <- so |>
  import_vdj(vdj_dirs, define_clonotypes = "cdr3_gene")
```

<br>

## Calculating gene usage

The `calc_gene_usage()` function will calculate the number of cells ('freq') and percentage of cells ('pct') with each gene in the `data_cols` column(s). The 'n_cells' column shows the total number of cells used for calculating percentages. By default these results are added to the object meta.data,
to return a data.frame set `return_df` to `TRUE`.

```{r}
so |>
  calc_gene_usage(
    data_cols = "v_gene",
    return_df = TRUE
  )
```

To perform gene usage calculations separately for cell clusters (or samples), provide the meta.data column containing cluster labels to the `cluster_col` argument. Here we see that the MD4 samples almost exclusively use a single V segment (IGKV5-43), which is expected since MD4 B cells are monoclonal.

```{r}
so |>
  calc_gene_usage(
    data_cols   = "v_gene",
    cluster_col = "sample",
    return_df   = TRUE
  )
```

To only perform calculations for a specific chain, use the `chain` argument. In this example we are only returning results for the IGK chain. Here we see some values in the 'v_gene' column labeled as 'None', this shows the number of cells that did not have a V gene segment identified.

```{r}
so |>
  calc_gene_usage(
    data_cols   = "v_gene",
    cluster_col = "sample",
    chain       = "IGK",
    return_df   = TRUE
  )
```

If two columns are provided to the `data_cols` argument, the number of cells containing each combination of genes is returned.

```{r}
so |>
  calc_gene_usage(
    data_cols   = c("v_gene", "j_gene"),
    cluster_col = "sample",
    return_df   = TRUE
  )
```

<br>

## Plotting gene usage

The `plot_gene_usage()` function will summarize the frequency of each gene segment. By default if a single column is passed to the `data_cols` argument, a bargraph will be returned. The number of top genes to include in the plot can be specified with the `genes` argument.

```{r}
so |>
  plot_gene_usage(
    data_cols   = "v_gene",
    cluster_col = "orig.ident",
    genes       = 20
  )
```

By default, percentages are shown on the y-axis, to instead plot the frequency, set the `units` argument to 'frequency'.

```{r}
so |>
  plot_gene_usage(
    data_cols   = "v_gene",
    cluster_col = "orig.ident",
    units       = "frequency"
  )
```

Plot colors can be adjusted using the `plot_colors` argument. In addition, `plot_gene_usage()` returns a ggplot object that can be modified with ggplot2 functions such as `ggplot2::theme()`. Plots can be further adjusted by passing aesthetic parameters directly to ggplot2, e.g. `alpha`, `linetype`, `color`, etc.

```{r}
so |>
  plot_gene_usage(
    data_cols   = "v_gene",
    cluster_col = "orig.ident",
    plot_colors = c(BL6 = "#3288BD", MD4 = "#D53E4F"),
    
    color = "black",  # parameters to pass to ggplot2
    alpha = 0.7
  ) +
  theme(axis.text.x = element_text(angle = 90))
```

If two columns are passed to the `data_cols` argument, a heatmap will be generated summarizing the usage of different pairs of segments. If a column is provided to the `cluster_col` argument, a separate heatmap will be generated for each cluster.

In this example we are plotting the frequency that different heavy chain V and J segments appear together. 

```{r, fig.height = 2.5}
so |>
  plot_gene_usage(
    data_cols   = c("v_gene", "j_gene"),
    cluster_col = "orig.ident",
    chain       = "IGH",
    genes       = 15
  )
```

The paired gene usage for two chains can also be plotted using `plot_gene_pairs()`. In this example we are plotting the frequency that different heavy and light chain V segments appear together.

```{r}
so |>
  plot_gene_pairs(
    data_col    = "v_gene",
    chains      = c("IGH", "IGK"),
    cluster_col = "orig.ident",
    genes       = 12
  )
```

<br>

## Circos plot

A circos plot can be created by setting the `method` argument to 'circos'. This plot will summarize the number of cells containing different gene pairs, which is shown as the axis labels for each sample. This requires the circlize package to be installed.

In this example, we are summarizing the segment usage for the entire dataset (BL6 and MD4 cells combined). The `cluster_col` argument can be used to create a separate plot for each sample. Labels can be rotated to eliminate overlapping text using the `rotate_labels` argument.

```{r, fig.width = 6, fig.height = 5}
so |>
  plot_gene_usage(
    data_cols     = c("v_gene", "j_gene"),
    method        = "circos",
    genes         = 6,
    rotate_labels = TRUE
  )
```

Plot colors can be modified using the `plot_colors` argument, additional parameters can be passed directly to `circlize::chordDiagram()`. In this example we add a border around the links and scale the plot so each sample is the same width.

```{r, fig.width = 6, fig.height = 5}
so |>
  plot_gene_usage(
    data_cols   = c("v_gene", "j_gene"),
    method      = "circos",
    genes       = 8,
    plot_colors = brewer.pal(10, "Spectral"),
    
    link.border = "black",  # parameters to pass to chordDiagram()
    scale       = TRUE
  )
```

Gene segment usage can be plotted separately for cell clusters (or samples) using the `cluster_col` argument. The number of rows used to arrange plots can be modified using the `panel_nrow` argument.

```{r, fig.width = 7, fig.height = 5}
so |>
  plot_gene_usage(
    data_cols   = c("v_gene", "j_gene"),
    method      = "circos",
    cluster_col = "sample",
    genes       = 5,
    panel_nrow  = 2,
    scale       = TRUE
  )
```

<br>

## Session info

```{r, echo = FALSE}
sessionInfo()
```
